Describe and Explore: Using feature-based time series clustering to identify meaningful structures in intensive longitudinal data

### 2023-01-08 manuscript.tex
:sparkles: self-similarity sect

To understand emotion dynamics, \citet{kuppens2017} originally proposed four features: (1) within-person variability, (2) co-variance or ICC, (3) inertia or autocorrelation, and (4) cross-lagged correlations. These features were then extended by \citet{krone2018}, adding (5) innovation variance, and (6) mean intensity. \citet{krone2018} even built a parametric model to tentatively cluster study participants. From a slightly different perspective \citet{dejonckheere2019} later added three additional features for psychological time series: (7) instability (8) interdependence (i.e., network density), and (9) diversity (i.e., Gini coefficient)\footnote{It should be noted that also within the psychological literature alternative summaries have been proposed that for example include measurement distribution, nonlinear developments, or categorical states. As an example, \citet{kiwuwa-muyingo2011} proposed to extract clinically meaningful states for medical adherence data and suggests these states as meaningful time series features.}.

@@

\paragraph{Variability} Variability, also referred to as dispersion or spread, captures to the degree to which a set of data differs from the central tendency \citep{weisberg1992}. In time series analyses, variability is important because it provides information about the distribution and diversity of the data. "Are people settled in their attitudes towards migrants or do they fluctuate over time?'' (i.e., difference in variability). 
There are several measures of variability, including the range, standard deviation, and variance (Hopkins, 2020). The appropriate measure to use depends on the specific characteristics of the data being analyzed. parametric: standard deviation, robust: median absolute deviation.

``Capturing variability may not just be helpful in identifying a worse psychological state or an inadequate emotion regulation strategy. It may also predict future course of symptoms. Affective variability at baseline predicted the recurrence of depressive symptoms over a three-year period in remitted depressed patients \citep{timm2017}. Similarly, elevated temporal auto-correlation and variance, as well as increased correlation between emotions, increased the probability for an upcoming shift between a normal and a depressed state \citep{vandeleemput2014}. ''\citep{myin-germeys2018}

EWS: anomalous variance \citep{wichers2019}

increase in variance \citep{scheffer2009}

\paragraph{(In)stability} Stability captures the average change between two consecutive measurements. While instability is conceptually related to the variability feature, variability does not take into account temporal dependency, whereas instability looks at the `jumpy-ness' of the data over time. In other words, instability reflects the fluctuation or inconsistency in a time series data over time, while variability reflects the range or diversity of values in a time series data \citep{trull2008}. For example, if a person has rapid and extreme changes in mood their mood is highly unstable, while if a person's mood responses span a wide range over the entire study period, their mood is highly variable \citep{jahng2008}. Common research questions with psychological time series might, thus, include: ``Do changes tend to be slow and gradual or fast and abrupt?" Mathematically, (in)stability measurements often look at the average change between two consecutive measurements \citep[e.g., mean absolute change; e.g.,][]{ebner-priemer2009,barandas2020}.

\paragraph{Self-similarity} Self-similarity describes how much a measurement carries over to future measurements. There are two main ways in which experiences tend to be connected to future experiences --- resistance to change and seasonality. Both forms of self-similarity relate to different types of research questions.

Resistance to change is the self-similarity of a measurement with its next measurement and is often referred to as inertia \citep{REF}. ``Do patients stay in a depressed mood for several measurements?'' 
characterize psychological maladjustment \citep{kuppens2010} and can even predict the initial onset of depression \citep{kuppens2012}.
Mathematically, inertia is often measured using autocorrelations or autoregressions with a lag-1.

Seasonality is the self-similarity of a measurement that returns in a cyclical manner and is often also referred to as periodicity \citep{REF}. ``Do participants drink more alcohol on Fridays?''. Both forms of self-similarity are often measured using autocorrelations with either a lag-1 (inertia) or the lag of the seasonality (e.g., seven days). Periodicity can additionally be captured using wave patterns \citep[e.g., fourier or wavelet transformation; e.g.,][]{mayor2022}.

>>>>

To understand emotion dynamics, \citet{kuppens2017} originally proposed four features: (1) within-person variability, (2) co-variance or ICC, (3) inertia or autocorrelation, and (4) cross-lagged correlations. These features were then extended by \citet{krone2018}, adding (5) innovation variance, and (6) mean intensity. \citet{krone2018} even built a parametric model to tentatively cluster study participants. From a slightly different perspective \citet{dejonckheere2019} later added three additional features for psychological time series: (7) instability (8) interdependence (i.e., network density), and (9) diversity \citep[i.e., Gini coefficient; also see][]{wendt2020}\footnote{It should be noted that also within the psychological literature alternative summaries have been proposed that for example include measurement distribution, nonlinear developments, or categorical states. As an example, \citet{kiwuwa-muyingo2011} proposed to extract clinically meaningful states for medical adherence data and suggests these states as meaningful time series features.}.

@@

\paragraph{Variability} Variability captures the degree to which a set of data differs from the central tendency and is sometimes also referred to as the dispersion or spread of the data \citep{weisberg1992}. In time series analyses, variability is conceptually important because it provides information about the distribution and diversity of the data. Especially within-person changes in variability have recently received increasing attention because they can be a potent early warning signal in psychological time series \citep{helmich2020a, vandeleemput2014}. A marked within-person increase in variance has, for example, been found to precede crucial transitions --- often considered to be part of a phenonenon know as critical slowing down \citep[i.e., anomalous variances; e.g.,][]{scheffer2009, wichers2019}.

But also more global person-level differences in time series variability have been found to be crucial and relate to important psychological questions. Person-specific differences in variability have, for example, been found to be indicative of worse psychological states \citep{myin-germeys2018}. And, similarly, person-level differences have also been associated with higher levels of psycho-pathological recurrences among depression patients \citep{timm2017}. But also more broadly speaking, psychological researchers and practitioners are theoretical-empirically often interested in person-level differences in variability. Researchers on polarization and radicalization might for example ask: "Are people settled in their attitudes towards migrants or vary across the measurement period?''

Methodologically, there are several measures of variability, including the range, standard deviation, and variance. The appropriate measure to use depends on the specific characteristics of the data being analyzed. For normally distributed data variances and standard deviation are often preferred, whereas data with outliers or nonparametric distributions are often better described with robust variability measures such as the median absolute deviation \citep{weisberg1992}.

\paragraph{(In)stability} Instability captures the average change between two consecutive measurements \citep{ebner-priemer2009}. While instability is conceptually related to the variability feature, variability does not take into account temporal dependency, whereas instability looks at the `jumpy-ness' of the data over time. In other words, instability reflects the fluctuation or inconsistency in a time series data over time, while variability reflects the range or diversity of values in a time series data \citep{trull2008}. For example, if a person has rapid and extreme changes in mood their mood is highly unstable, while if a person's mood responses span a wide range over the entire study period, their mood is highly variable \citep{jahng2008}. Within psychological time series, instability measurements have especially been important in the research of borderline personality disorder \citep{trull2008} and suicidality \citep{kivela2022}, but also in understanding early warning signals more generally \citep{wichers2019}. Conceptually, the instability feature, thus, relates to a broad range of research questions, including: ``Do changes tend to be slow and gradual or fast and abrupt?" Methodologically, (in)stability measurements often look at the average change between two consecutive measurements, such as the mean absolute change \citep[e.g.,][]{ebner-priemer2009,barandas2020}.

\paragraph{Self-similarity} Self-similarity in time series data refers to the property of a time series to exhibit similar patterns of behavior over different time scales \citep{dmello2021}. That is, self-similarity describes how much a measurement carries over to future measurements. There are two main ways in which experiences tend to be connected to future experiences in psychological time series --- resistance to change and seasonality. Both forms of self-similarity relate to different types of research questions.

Resistance to change is the self-similarity of a measurement with its next measurement and is often referred to as inertia \citep{kuppens2010}. If inertia is high a development tends to stay in a certain state or move towards a certain direction. Because high inertia is resistant to change, in emotion dynamics high inertia has been found to be indicative of underreactive systems and to be characteristic of psychological maladjustment \citep{kuppens2010}. In a similar vein, high inertia at baseline was even predictive of the initial onset of depression \citep{kuppens2012}. Conceptually, inertia is more broadly connected to research questions such as: ``Do patients stay in a depressed mood for several measurements?'' One way to measure inertia in time series data is through the use of autocorrelations or autoregressions with a lag-1, which look at the correlation between a given time series and itself at the previous measurement occasion. High autocorrelation values can indicate high levels of inertia, while low autocorrelation values may indicate a more unpredictable or volatile time series \citep{dejonckheere2019}.

Seasonality is the self-similarity of a measurement that returns in a cyclical manner and is often also referred to as periodicity \citep{gregson1983}. Periodicity in psychological time series data, thus, refers to the presence of recurring patterns at regular intervals. In psychological time series, these cyclical patterns are not only dangerous for models that assume stationarity \citep{beal2015} but can be interesting psychological properties in their own regard \citep{epskamp2018, schmittmann2013}. Humans generally live in seasonal environments, including, for example, exam periods at universities \citep{fuller2003}. But also more internally, personality manifestations, for example, have long been assumed to develop cyclically \citep{cattell1957}. The associated research questions can be driven by an understanding of external seasonality factors such as: ``Do participants drink more alcohol on Fridays?'' But research questions can also address periodic patterns as they may be the result of underlying psychological processes: ``Does stress increase with fatigue cycles?'' Similar to inertia, periodicity is often measured using autocorrelations but then with the lag of the seasonality or cycle (e.g., daily, seven days, one month, one semester). Periodicity can additionally be captured using spectral analysis or decomposition, such as with wave patterns \citep[e.g., fourier or wavelet transformation; e.g.,][]{mayor2022}.
